The Array of Long-Baseline Antennas for Taking Radio Observations from the Sub-Antarctic

### 2020-07-20 albatros_paper.tex
HCC added more text in observations and conclusions

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.8\linewidth]{Figures/baseband_direct_auto_compare.jpg}
    \caption{Placeholder figure for comparison of auto-spectra from
      baseband (4-bit), and directly computed/averaged by the SNAP.}
    \label{Fig:baseband_direct_auto}
  \end{center}
\end{figure}

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.8\linewidth]{Figures/baseband_direct_phase_compare.jpg}
    \caption{Placeholder figure for comparison of cross-spectrum phase from
      baseband (4-bit), and directly computed/averaged by the SNAP.}
    \label{Fig:baseband_direct_phase}
  \end{center}
\end{figure}

\section{Preliminary observations}

The primary goal of the two-element pathfinder described in
\S\ref{s:2elem} was to qualitatively assess the level of RFI and
ionospheric contamination at low frequenices.
\autoref{Fig:albatros2_waterfalls} illustrates auto- and cross-spectra
from one of the co-aligned polarization pairs from the two antennas.
The data are shown over a three-day period (June 18--21, 2018), and no
calibration or cuts have been applied.  The rising and setting of the
Galaxy is visible in all of the spectrum magnitudes, and repeatable
fringes are visible in the phases of the cross-spectra.  The signal
drop-off below $\lesssim30$~MHz in the autospectra is caused largely
by the combined response of the LWA antenna and FEE.  Although the LWA
antenna is not optimized for the lowest \albatros\ observing
frequencies, as described in \S\ref{s:antenna}, it is sufficient for
these initial, exploratory measurements.  The RFI lines at
$\lesssim20$~MHz arise primarily from shortwave radio transmission
reflecting off the ionosphere.  During the night, when the ionospheric
plasma cutoff frequency drops, the maximum frequency of the RFI
contamination falls to $\sim10$~MHz.

@@

We have been unable to retrieve data from the single autonomous
station pathfinder on Marion because of COVID-19 and limited bandwidth
to the island, although we have confirmed that the solar power system
has successfully met the power demands of the antenna station.  To
demonstrate baseband writing capability in the absence of Marion data,
we instead present a short observation that was taken by a separate
\albatros\ pathfinder station that operated from the McGill Arctic
Research Station (MARS; \ang{79;26;}N, \ang{90;46;}W) during
July~2019.  The RF signal chain of the MARS installation had the same
architecture as that shown in \autoref{Fig:albatros1_schem}, but the
power was supplied by manually charged batteries, rather than a solar
power system.  The MARS RFI environment is somewhat noisier than that
of Marion Island, and because the data were taken during the height of
Arctic summer, the ionospheric plasma cutoff frequency is higher than
the typical cutoff reflected in the Marion data.

\autoref{Fig:baseband_direct_auto} shows autospectra for one
polarization of the \albatros\ pathfinder installed at MARS, comparing
the accumulated spectra that are directly computed by the SNAP board
against spectra that are calculated from baseband data with 4-bit
quantization.  (Autospectrum information is not preserved with 1-bit
quantization.)  Baseband data were recorded over two frequency
windows, 5.3--12.6~MHz and 12.6--20.0~MHz, at separate times, and the
corresponding accumulated spectra from the SNAP board are shown for
the same frequency and time ranges.  The waterfall plots have the same
61-kHz frequency resolution, set by the resolution of the spectra
computed by the SNAP board.  The overplotted mean spectra are shown
with finer resolution (\attention{numbers go here}) from the baseband
data to illustrate the additional information that is not retained in
the accumulated spectra.  For these mean spectrum plots, the baseband
data are processed by
inverting\footnote{\url{https://github.com/jrs65/pfb-inverse}} the
polyphase filter bank to recover time-domain data, and the forward
polyphase filter bank is then applied to concatenated timestreams of
an appropriate total length to set the desired frequency resolution.
The spectra computed from baseband qualitatively agree with those
directly accumulated by the SNAP board, up to small differences caused
by the 4-bit quantization.

\attention{more words to come here about phase plots}

\section{Conclusions}

\attention{text goes here}

% We present the basics of 1-bit correlation here, and show that even in
% the high-signal limit, 1-bit correlation remains viable.

% discuss design changes for final stations: larger banks of hard
% drives, possible wind power, etc.

>>>>

\section{Preliminary observations}\label{s:obs}

The primary goal of the two-element pathfinder described in
\S\ref{s:2elem} was to qualitatively assess the level of RFI and
ionospheric contamination at low frequencies.
\autoref{Fig:albatros2_waterfalls} illustrates auto- and cross-spectra
from one of the co-aligned polarization pairs from the two antennas.
The data are shown over a three-day period (June 18--21, 2018), and no
calibration or cuts have been applied.  The rising and setting of the
Galaxy is visible in all of the spectrum magnitudes, and repeatable
fringes are visible in the phases of the cross-spectra.  The signal
drop-off below $\lesssim30$~MHz in the autospectra is caused largely
by the combined response of the LWA antenna and FEE.  Although the LWA
antenna is not optimized for the lowest \albatros\ observing
frequencies, as described in \S\ref{s:antenna}, it is sufficient for
these initial, exploratory measurements.  The RFI lines at
$\lesssim20$~MHz arise primarily from shortwave radio transmission
reflecting off the ionosphere.  During the night, when the ionospheric
plasma cutoff frequency drops, the maximum frequency of the RFI
contamination falls to $\sim10$~MHz.

@@

We have been unable to retrieve data from the single autonomous
station pathfinder on Marion because of COVID-19 and limited bandwidth
to the island, although we have confirmed that the solar power system
has successfully met the power demands of the antenna station.  To
demonstrate baseband writing capability in the absence of Marion data,
we instead present a short observation that was taken by a separate
\albatros\ pathfinder station that operated from the McGill Arctic
Research Station (MARS; \ang{79;26;}N, \ang{90;46;}W) on Axel Heiberg
Island, Nunavut, during July~2019.  The RF signal chain of the MARS
installation had the same architecture as that shown in
\autoref{Fig:albatros1_schem}, but the power was supplied by manually
charged batteries, rather than a solar power system.  The MARS RFI
environment is somewhat noisier than that of Marion Island, and
because the data were taken during the height of Arctic summer, the
ionospheric plasma cutoff frequency is higher than the typical cutoff
reflected in the Marion data.  Despite the differences in the power
configuration and RFI environment, the MARS installation is suitable
for illustrating baseband data writing.

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.8\linewidth]{Figures/baseband_direct_auto_compare.jpg}
    \caption{\attention{Placeholder figure} Autospectra for one
      polarization from an \albatros\ pathfinder at MARS (see text for
      details).  The top row shows spectra that are accumulated
      directly by the SNAP board, and the middle row shows spectra
      computed from baseband data with 4-bit quantization.  The two
      frequency windows, 5.3--12.6~MHz and 12.6--20.0~MHz, are
      recorded at two separate times, and all waterfall plots are
      shown with 61-kHz resolution.  The bottom row compares the mean
      autospectra, with finer, \attention{numbers here} resolution of
      the baseband spectra.}
    \label{Fig:baseband_direct_auto}
  \end{center}
\end{figure}

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.8\linewidth]{Figures/baseband_direct_phase_compare.jpg}
    \caption{\attention{Placeholder figure} Phases from the
      cross-correlation between two orthogonal polarizations from an
      \albatros\ pathfinder at MARS (see text for details).  The top
      row shows cross-spectrum phases that are accumulated directly by
      the SNAP board, and the bottom row shows phases computed from
      baseband data with 4-bit quantization.  The two frequency
      windows, 5.3--12.6~MHz and 12.6--20.0~MHz, are recorded at two
      separate times, and all waterfall plots are shown with 61-kHz
      resolution.}
    \label{Fig:baseband_direct_phase}
  \end{center}
\end{figure}

\autoref{Fig:baseband_direct_auto} shows autospectra for one
polarization of the \albatros\ pathfinder installed at MARS, comparing
the accumulated spectra that are directly computed by the SNAP board
against spectra that are calculated from baseband data with 4-bit
quantization.  (Autospectrum information is not preserved with 1-bit
quantization.)  Baseband data were recorded over two frequency
windows, 5.3--12.6~MHz and 12.6--20.0~MHz, at separate times, and the
corresponding accumulated spectra from the SNAP board are shown for
the same frequency and time ranges.  All of the waterfall plots have
the same 61-kHz frequency resolution, set by the resolution of the
spectra computed by the SNAP board.  The overplotted mean spectra are
shown with finer resolution (\attention{numbers go here}) from the
baseband data to illustrate the additional information that is not
available in the accumulated spectra.  For these mean spectrum plots,
the baseband data are processed by
inverting\footnote{\url{https://github.com/jrs65/pfb-inverse}} the
polyphase filter bank to recover time-domain data, and the
forward-direction polyphase filter bank is then applied to
concatenated timestreams of an appropriate total length to set the
desired frequency resolution.  \autoref{Fig:baseband_direct_phase}
shows waterfall plots of phases computed from the cross-correlation
between the two orthogonal polarizations of the \albatros\ pathfinder
at MARS, comparing the directly accumulated and baseband data.  For
both the autospectra and phases, the baseband data qualitatively agree
with the data directly accumulated by the SNAP board, up to small
differences caused by the 4-bit quantization.

\section{Summary and future outlook}

We have presented the design of \albatros, a new interferometer that
will image the radio sky at $\lesssim30$~MHz using an array of
autonomous antenna stations installed on Marion Island.  With a
two-element, directly correlated pathfinder, we have demonstrated that
clear, repeatable sky signal is visible from Marion down to
$\lesssim10$~MHz with no data processing or cuts.  We have constructed
the first prototype autonomous \albatros\ station powered by solar
panels, and we have successfully tested the electronics and software
for recording baseband data.

With the proof of concept demonstrated from the pathfinder instruments
presented here, we plan to improve the design of the future
\albatros\ stations that will be installed at the coastal huts on
Marion.  The LWA antenna and FEE are currently not optimized for the
lowest \albatros\ observing frequencies, and we are investigating
possible modifications to the antenna and FEE designs to improve the
low frequency response, as well as building in calibration circuitry.
Because the future \albatros\ stations will be located farther from
the base and will be more difficult to access on a regular basis, each
station will require larger total hard drive volume to store baseband
data over extended periods of time.  We are developing a custom
low-power hard drive bank with $>100$~TB total storage capacity that
will employ a USB multiplexer to select and power only one hard drive
at a time.  The autonomous station pathfinder presented here uses
solar panels to charge the batteries, and we are investigating the use
of small wind turbines as a possible alternative for future stations.

In addition to Marion Island, we are also planning on installing a
second \albatros\ array at MARS in the high Arctic.  As described in
\S\ref{s:obs}, a pathfinder antenna was installed in July 2019 and
observed for about three weeks to assess the RFI environment and
ionospheric conditions.  When fully operational, the Marion and MARS
\albatros\ installations will provide new views of the low-frequency
sky across both hemispheres.
